Best Practices in Latent Class Analysis using Free Open Source Software

### 2022-10-16 manuscript.Rmd
Update 16 October
Changed title and short title. Adjusted abstract, Added keywords. Made the use of "LCA" and "mixture models" more consistent based on whether we refer to the whole group of LCA techniques, or a mixture subset.  Minor adjustment to wording throughout the paper. Shortened lines throughout. Capitalized titles and made them uniform. Adjusted section-subsection-subsubsection depth (# vs ## vs ###).  Removed irrelevant comments. Wrote best practices in visualization (##Visualization).  Added the tutorial (now set to eval = F for knitting, however code chunks do not show when using

---
title             : "Best Practices in Mixture Modeling using Free Open Source Software"
shorttitle        : "BEST PRACTICES MIXTURE MODELING"

@@

abstract: |
  Latent class analysis is a popular technique for identifying groups in data based
  on a parametric model. Examples of this technique are known as mixture models, latent profile
  analysis, latent class analysis, growth mixture modeling, and latent class
  growth analysis. Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture
  models in the `tidySEM` R-package. These functions rely on estimation using
  the `OpenMx` R-package.

keywords          : "keywords"
wordcount         : "X"
bibliography      : "mixture.bib"
figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
mask              : no

class             : "man"
output            : papaja::apa6_pdf
---

@@

Latent class analysis (LCA) is an umbrella term that refers to a number of techniques for estimating unobserved group membership
based on a parametric model of one or more observed indicators of group membership.
This method has become quite popular across scientific fields, and under a number of different names;
most notably (finite Gaussian) mixture modeling and latent profile analysis.
@vermunt_jk_latent_2004 defined it more generally as virtually any statistical model where “some of the parameters […] differ across unobserved subgroups”.

Despite the popularity of the method, there is a lack of standards for estimating and reporting latent class analyses.
While @van_de_schoot_grolts-checklist_2017 developed reporting guidelines for a specific subcategory of LCA known as latent growth models,
general reporting guidelines for latent class analysis are still lacking.
This complicates manuscript review and assessment of the quality of published studies,
and introduces a risk of misapplications of the technique.
The present paper seeks to address this gap in the literature by suggesting updated guidelines for estimation and reporting on latent class analysis, based on current best practice.
Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples, 
this paper exclusively relies on free, open source software for latent class analysis in R.
Our goal is to make best-practices in latent class analysis widely accessible.

## Defining latent class analysis

Latent class analysis can be understood as a method for estimating unobserved groups
based on a parametric model of observed indicators of group membership.
The concept of latent class analysis can be understood in different ways.
Generally speaking, a mixture model assumes that the study population
is composed of $K$ subpopulations or classes.
It further assumes that the observed data are a mixture of data generated by class-specific models.
The simplest univariate "model" is a normal distribution, which can be described with two parameters:
the mean and the variance.
Commonly, the same model is estimated across all classes, but with different parameters for each class (i.e., class-specific means and variances).
Mixture modeling then estimates both the parameters for each class, 
and the probability that an individual belongs to each class.

@@

A mixture model is like confirmatory factor analysis, except that the continuous latent variable is substituted with a categorical latent variable.
One difference between the two techniques is that factor analysis can be considered as a way to group observed *variables* into latent constructs, with factor loadings indicating which items belong are most indicative of a construct.
By contrast, mixture modeling groups *individuals* into classes [see @nylund-gibson_ten_2018].
In line with this distinction, latent class analysis is sometimes referred to as a "person-centered" technique, and factor analysis as a "variable-centered" technique.

When the focus is on the model parameters in each group,
then LCA can be thought of as similar to a multi-group structural equation model.
The main distinction is that group membership is not known a-priori,
but is instead estimated - with measurement error - based on the data.
Whereas in a multi-group model, the data are split by group and treated as independent samples,
in a mixture model, all cases contribute to the estimation of all parameters in all groups.
The relative contribution of each case to the parameters of each group is determined by that case's posterior probability of belonging to that group.

When the focus is on each individual's estimated class membership,
latent class analysis can be thought of as a type of clustering algorithm.
In line with this perspective, mixture modeling is sometimes described as "model-based
clustering" [@hennig_handbook_2015; @scrucca_mclust_2016].
Many clustering algorithms apply some recursive splitting algorithm to the data.
By contrast "model-based" clustering refers to the fact that LCA 
estimates cluster membership based on a parametric model.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.

Finally, in the context of machine learning, LCA can be considered as an
*unsupervised classification* problem [@figueiredo_unsupervised_2002].
The term *unsupervised* refers to the fact that the outcome variable, true class membership,
is not known, and the term *classification* refers to the fact that the algorithm is predicting a categorical outcome -- class membership.

## A taxonomy of latent class analyses

In this paper, we use the term latent class analysis to refer to techniques that estimate latent class membership based on a parametric model of observed indicators.
From a historical perspective, the term latent class analysis was initially conceived to refer to analyses with categorical (binary) indicators [@vermunt_jk_latent_2004].
Nowadays, there are a number of related techniques, known by distinct names, that serve a similar purpose.
The term "latent class analysis" seems most appropriate as an umbrella term for this broader class of models,
as it only refers to the purpose of the analysis, and does not imply restrictions to the model used, or the level of measurement of the indicators.
Given the abundance of terms in use for closely related classes of models, we will provide a rudimentary taxonomy of latent class analyses.

One common type of LCA is the *finite Gaussian mixture model*;
a univariate analysis where the observed distribution of a single variable is assumed to result from a mixture of a known number of Gaussian (normal) distributions.
The parameters of a finite Gaussian mixture model are the means and variances of these underlying normal distributions.
The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
In the multivatiate case, with more than one indicator variable,
the parameters of a mixture model are the means, variances, and covariances between the indicators (which can be standardized to obtain correlations).
These parameters can be estimated freely, or set to be constrained across classes.

The technique known as *latent profile analysis (LPA)* is a special case of the mixture model,
which assumes conditional independence of the indicators.
Conditional independence means that, after class membership is accounted for,
the covariances/correlations between indicators are assumed to be zero.
This can be conceived of as a restricted mixture model with covariances fixed to zero.
In some cases, such constraints will be inappropriate;
for example, when the cohesion between indicators is expected to differ between classes.
As an example, a mixture model analysis of ocean plastic particles found two classes of particles based on length and width:
a class of smaller particles with a high correlation between length and width, meaning that these particles were approximately round or square in shape, and a class of larger particles with a low correlation between length and width, meaning that these particles were heterogenous in shape.
From a theoretical perspective, this makes sense, because the smaller particles have been ground down to a more uniform shape by the elements.
<!-- DA: what is the source for plastic particles? -->

It is also possible to estimate a mixture model based on latent indicators.
This means that, within each class, one or more continuous latent variables are estimated based on the observed indicators.
Categorical latent variable membership is then estimated based on these continuous latent variables.
A common application of this approach is in longitudinal research, where the indicators reflect one construct assessed at different time points.
Examples of this approach include *growth mixture models* (GMM) and *latent class growth analyses* (LCGA).
These techniques estimate a latent growth model to describe individual trajectories over time.
The growth mixture model is a latent class model where the parameters that indicate class membership 
are the intercepts and variances (and typically covariances) of the latent growth variables, e.g., a latent intercept and slope.
This technique assumes that individuals within a class can have heterogenous trajectories.
If the variance of the growth parameters is fixed to zero, it is known as a latent class growth analysis.
This latter approach assumes that all individuals within a class share the same identical trajectory,
and that any variance in the indicators not explained by the class-specific latent trajectories is due to residual error variance.

@@

This paper will primarily focus on mixture models and special cases thereof,
although most of the suggested guidelines are applicable to all latent class analyses.

## Use cases for latent class analysis

<!-- CJ: It's good but for practical purposes it's preferable -->
<!-- to use examples from Caspar's published papers. -->
<!-- Don't spend too much time on this, -->
<!-- Caspar will check if I can replace them with any of my papers. -->

There are several use cases for which latent class analyses are suitable.
One example is to test a theory that postulates the existence of a categorical latent variable.
For example, *identity status theory* posits that, at any given point in time, adolescents reside in one of four identity statuses.
Latent class analysis can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment).
If results indicate that the data are better described by a different number of classes,
or that the four-class solution does not correspond to the predicted pattern of responses on the indicators, then the theory may be called into question.

Another use case is unsupervised learning;
when the goal is to restore unobserved class membership based on observed indicators,
or to classify individuals.
For example, a mixture model can be used as a diagnostic aid when several clinical indicators can be used to distinguish between a fixed number of physical [@baughman_mixture_2006] or mental [@wu_abuse_2011] health problems.
The example of shoe size is a rudimentary illustration of this type of application.

@@

With LCA, our goal could be to inductively identify the number of classes. 
For instance, if we believe that a variable represents a group, 
but don't know how many groups there are, LCA may be an appropriate technique to answer this question.
For example, @hopfer_social_2014 studied the substance use, sexual behavior, and mental health status of urban population in Winnipeg, Canada. 
The underlying assumption was that there were different risk profiles,
but their number was not known.
From a collection of indicators, the LCA provided evidence that there may be four distinct risk profiles in the Winnipeg area.

Another application of LCA is to classify individuals.
In a peer harassment study, @giang_using_2008 used latent class analysis to classify over 2,000 sixth grade students into aggressor and victim latent classes.
The five-class solution comprised of classes of victims, aggressors, and socially adjusted students. For instance, it revealed that there were two types of victims: highly-victimized aggressive-victims and highly-aggressive aggressive-victims.

LCA is also appropriate when we wish to identify indicators that capture classes well.
High quality indicators are strongly related to the latent variable
and lead to good class separation.
This relationship of high quality indicators to the latent variable
is reflected in very high or very low conditional response probabilities.
<!-- As a rule of thumb, high indicator quality can be indicated  -->
<!-- by conditional response probabilities above .9 and below .1.  -->
<!-- Moderate indicator quality can be indicated by probabilities above .8 and below .2, -->
<!-- and low indicator quality would be above .7 and below .3. <- DA's writing -->
<!-- This quality rule of thumb is from @geiser_is_2014 where they state that -->
<!-- "Indicators in high quality conditions were generated to either have CRPs (of endorsing the second category) of 0.9 or 0.1, while moderate quality conditions had CRPs of 0.8 and 0.2, and low quality indicators were generated at 0.7 or 0.3, following Collins and Wugalter’s (1992) strong and weak measurement strength conditions." direct quote from @geiser_is_2014 -->
For a simulation study exploring the effects of indicator quality on LCA, see @geiser_is_2014.
Therefore, one could use conditional response probabilities for each item to assess its quality
with regards to how well it helps separate the latent classes.
From this, a theory about the selection of indicators could be informed.

<!-- "However, Galindo-Garre and Vermunt (2006) found that indicators with high population values (i.e., conditional response probabilities close to one) were more likely to produce boundary estimates in LCA applications, even though this effect decreased with larger N. Still, there is evidence that using high quality indicators is generally beneficial, at least in the context of structural equation models with continuous latent variables (Marsh et al., 1998). Having more high quality indicators can stabilize the estimation by increasing the information available to estimate latent variable parameters. In this vein, Collins and Wugalter (1992) speculated that having sufficiently high quality indicators may compensate for having few indicators and may aid parameter recovery in LTA models." direct quote from @geiser_is_2014 -->

An extension of LCA is that containing covariates which can be used to predict class membership.
In this approach, we not only model the latent class variable based on indicators,
but we also relate the class membership to other explanatory variables (@vermunt_latent_2017).
An example of using covariates comes from @nozadi_moderating_2016 
who applied LCA to identify the probability of children's membership to an anxiety class.
The authors tested several covariates including children's age, sex, and accuracy scores.
Age and sex were not found to be related to the children's latent class membership,
hence these covariates were excluded from the analysis.
Accuracy scores were related to probabilities of being in anxiety and attention-
anxiety classes and therefore this covariate was kept as a valuable predictor.

@@

<!-- from @macgregor_symptom_2021 "Incorporating LCA with the symptom checklist was a novel method of dimensionality -->
<!-- reduction, and the identified symptom profiles provided more granularity than examining individual symptoms -->
<!-- alone. There are also limitations that warrant mention. At times, LCA can be prone to classification errors, but in the present analysis, the model entropy value above 0.80 indicated good separation of the classes, and the MeanPP values showed a low rate of misclassification." -->
<!-- [...] -->
<!-- [listing the five classes] "(1) 50.4% were low -->
<!-- morbidity (i.e., no or very few symptoms reported), (2) -->
<!-- 14.0% musculoskeletal (i.e., muscle aches; swollen, stiff or -->
<!-- painful joints; back pain), (3) 8.9% auditory (i.e., trouble -->
<!-- hearing; ringing in the ears), (4) 11.1% psycho-cognitive -->
<!-- (i.e., trouble concentrating, easily distracted; forgetful or -->
<!-- trouble remembering things; increased irritability), and (5) -->
<!-- 15.6% multimorbidity (i.e., multiple symptoms reported)." -->

<!-- CJ: (on the next paragraph)  -->
<!-- the multilevel LCA addresses violations of the assumptions of LCA. -->
<!-- That is not what this paragraph is about.  -->
<!-- This paragraph is about motivations for using LCA. -->
<!-- One motivation is if your data violates multivariate normality -->
<!-- because violating multivariate normality means that you have multiple groups -->
<!-- and you can use LCA to extract those. That is the only assumption  -->
<!-- we have to discuss here, we just need to explain it. -->
Finally, LCA can be used to deal with data which violate certain assumptions.
As discussed in the shoe size example, LCA can deal with violations of normality.
In fact, LCA assumes the population distribution in a non-normal mixture of $K$ normal distributions,
and it can discover the value of $K$, i.e. generate a $K$-class solution.

<!-- CJ: Include these use cases too: -->
<!-- Descriptive analysis; identify few prototypes based on many variables, e.g. DJA science funding -->
<!-- Inductively identify number of latent classes: I know it's a group, but not how many groups -->
<!-- Classify individuals, e.g. welzijnsmeter -->
<!-- Identify items (observed characteristics) that -->
<!-- indicate these classes well -->
<!-- Predict class membership from covariates -->
<!-- Predict outcomes from class membership -->
<!-- Dimension reduction -->
<!-- Deal with violation of assumptions, e.g. when data are clearly not normally distributed -->

<!-- In medical and psychological science, LPA can be useful when -->
<!-- considerable between-subject heterogeneity exists in scores on a range of variables and when -->
<!-- this variation cannot be explained by known, manifest variables (e.g., Wolfe 1970; Sterba -->
<!-- 2013). Here, LPA can help to identify or approximate possibly meaningful sub-grouping of -->
<!-- subjects that may help to better understand sample heterogeneity (Sterba 2013). -->
<!-- Generally, LPA works under the assumption that sample (residual) variance can be reduced -->
<!-- by assuming a categorical latent variable that effectively subdivides the sample into >=2 -->
<!-- subgroups that are more homogeneous in terms of their patterns of variable means and -->
<!-- (co)variances. In case an LPA model is found to fit a dataset well, it is often the case that -->
<!-- subjects in each class resemble each other relatively well in terms of their scores on the input -->
<!-- variables. Depending on the model configuration, the identified classes can show different -->
<!-- class-specific patterns of means and class-specific or class-varying variances. -->

<!--                                                1 -->
<!-- A downside is that these models are more complex (much more parameters to be estimated). -->
<!-- Using criteria, such as the Bayesian Information criterion (BIC) helps to find a model that -->
<!-- strikes a good balance between model-fit and model-complexity. When doing an LPA, most -->
<!-- applied researchers will usually be most interested in the differences and/or overlap between -->
<!-- the classes’ specific patterns of parameter estimates. These can be used to characterize the -->
<!-- classes and, possibly, provide clues about underlying mechanisms (Sterba 2013). -->

<!-- This tutorial -->
<!-- LPA is less widely used than other latent variable models and, possibly due to that, has -->
<!-- long been only available in specialized software packages such as Mplus. Luckily, ongoing -->
<!-- developments in many different scientific fields (e.g., ecology, econometrics) have yielded a -->
<!-- number of packages that also allow users to conduct LPA in the open-source R-platform. -->
<!-- However, the use of R does require experience and documentation of packages can be rather -->
<!-- limited or technical, making it less easily accessible for applied researchers. Therefore, this -->
<!-- tutorial is aimed to help applied researchers to get going with LPA in R, illustrating the use -->
<!-- of several packages and, for reference, providing a comparison of the results obtained in R -->
<!-- with results obtained with Mplus. -->

# Best practices

## In estimation

The best practices in estimation, as outlined in Table \@ref(tab:checkest),
are rooted in existing recommendations for best practices for estimating
specific sub-types of latent class analyses,
including latent class growth analysis [@van_de_schoot_grolts-checklist_2017]
and latent class analysis with ordinal indicators [e.g., @nylund-gibson_ten_2018].
These were generalized to be more relevant to all types of latent class analyses,
and updated to current best practices, as explained below.

```{r checkest, include=FALSE}
tab_check <- data.frame(Item = c(
  "Examine observed data",
  "Handling missing data",
  "Alternative model specifications",
  "Software",
  "Best practices in estimation",
  "Algorithm",
  "Class enumeration",
  "Model Fit Indices",
  "Classification Diagnostics",
  "Interpreting Class Solution",
  "Label switching",
  "Best Practices in Reporting",
  "Tutorial"

@@

### Examine observed data

Examining observed data is essential for any analysis
as it may reveal patterns and violations of assumptions that had not been considered prior to data collection.
Special attention should be paid to level of measurement of the indicators.
Finite Gaussian mixture models (including LPA) are only suitable for continuous variables.
Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions of mixture models [see @vermunt_k-means_2011].
Personal experience consulting on latent class analyses and moderating the `tidyLPA` Google group
suggest that the misapplication of mixture models to ordinal (e.g., Likert-type) indicators is the most common source of user error.
Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous [e.g., @norman_likert_2010],
this certainly does not imply that all methods are.
It is certainly unlikely that such ordinal variables can be treated as a *mixture* of multiple normal distributions.
The problem becomes egregious when the number of classes estimated equals or exceeds the number of categories; 
in this case, each class-specific mean could describe a single response category, 
and a class-specific variance component would be nonsensical.
In sum, Likert-type scales are rarely suitable for mixture modeling;
latent class analysis with ordinal indicators is more appropriate.

Relatedly, a recent publication claimed that an assumption of mixture models is that observed indicators are normally distributed [@spurk_latent_2020].
This is incorrect.
When the number of classes is greater than one, mixture models assume that the observed indicators are a mixture of multiple (multivariate) normal distributions.
In our shoe size example, it can be seen that the population distribution comprised of two normal distributions.
When examined visually, the population distribution is evidently bimodal. 
The Shapiro-Wilk normality test ($W = 0.971$, $p < 0.05$ ) 
rejects the null hypothesis that 
the sample comes from a normally distributed population.
Yet, this is a prototypical example of a mixed population distribution where LCA can discover latent groups.
If the population distribution were instead normal, 
there would be no classes to extract 
as the whole population would belong to a single class.

Extensive descriptive statistics (including the number of unique values, variance of categorical variables, and missingness; see next paragraph) can be obtained using the function `tidySEM::descriptives(data)`.
Note, however, that sample-level descriptive statistics are of limited value when the goal of a study is to identify sub-samples using latent class analysis.
Plots (density plots for continuous variables, and bar charts for categorical ones) may be more diagnostic.
Note that density plots can also aid in the choice of the number of classes, as further explained in the section on visualization.
Descriptive statistics and plots can be relegated to online supplements, provided that these are readily accessible [consider using a GitHub repository as a comprehensive public research archive, as explained in @van_lissa_worcs_2021].

### Handling missing data

>>>>

---
title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
shorttitle        : "BEST PRACTICES LCA"

@@

abstract: |
  Latent class analysis is a family of techniques for identifying groups in data based
  on a parametric model. Its examples include mixture models, latent profile
  analysis, latent class analysis with ordinal indicators, and latent class
  growth analysis. Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture
  models in the `tidySEM` R-package whose backend relies on
  the `OpenMx` R-package.

keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
wordcount         : "8206"
bibliography      : "mixture.bib"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no

class             : "man"
output            : papaja::apa6_pdf

---

@@

Latent class analysis (LCA) is an umbrella term 
that refers to a number of techniques for estimating unobserved group membership
based on a parametric model of one or more observed indicators of group membership.
The types of LCA have become quite popular across scientific fields,
most notably finite Gaussian mixture modeling and latent profile analysis.
@vermunt_jk_latent_2004 defined LCA more generally as virtually any statistical model 
where “some of the parameters […] differ across unobserved subgroups”.

Despite its popularity, there is a lack of standards for estimating and reporting LCA.
While @van_de_schoot_grolts-checklist_2017 developed reporting guidelines 
for a specific type of LCA known as latent growth models,
general reporting guidelines for all LCA are still lacking.
This complicates manuscript review and assessment of the quality of published studies,
and introduces a risk of misapplications of the technique.
The present paper seeks to address this gap in the literature 
by suggesting updated guidelines for estimation and reporting of LCA, 
based on current best practices.
Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples, 
this paper exclusively relies on free, open source software in R.
Our goal is to make best-practices in LCA widely accessible.

## Defining Latent Class Analysis

Latent class analysis is a group of methods for estimating unobserved groups
based on a parametric model of observed indicators of group membership.
The concept of LCA can be understood in different ways.
A mixture model assumes that the study population
is composed of $K$ subpopulations or classes.
It further assumes that the observed data are a mixture of data generated by class-specific models.
The simplest univariate model is a normal distribution, 
which can be described with two parameters: the mean and the variance.
Commonly, the same model is estimated across all classes, 
but with different parameters for each class (i.e., class-specific means and variances).
Mixture modeling then estimates both the parameters for each class, 
and the probability that an individual belongs to each class.

@@

A LCA is like confirmatory factor analysis, 
except that the continuous latent variable is substituted with a categorical latent variable.
One difference between the two techniques is that 
factor analysis can be considered as a way to group observed *variables* into latent constructs, 
and LCA groups *individuals* into classes.
In line with this distinction, LCA is sometimes referred to as a "person-centered" technique, 
and factor analysis as a "variable-centered" technique [@nylund-gibson_ten_2018].

When the focus is on the model parameters in each group,
LCA can be thought of as similar to a multi-group structural equation model.
The main distinction is that group membership is not known a-priori,
but is instead estimated -- with measurement error -- based on the data.
Whereas in a multi-group model, the data are split by group and treated as independent samples,
in LCA, all cases contribute to the estimation of all parameters in all groups.
The relative contribution of each case to the parameters of each group
is determined by that case's posterior probability of belonging to that group.

When the focus is on each individual's estimated class membership,
LCA can be thought of as a type of clustering algorithm.
In line with this perspective, LCA is sometimes described as "model-based
clustering" [@hennig_handbook_2015; @scrucca_mclust_2016].
Many clustering algorithms apply some recursive splitting algorithm to the data.
By contrast "model-based" clustering refers to the fact that LCA 
estimates cluster membership based on a parametric model.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.

Finally, in the context of machine learning, LCA can be considered as an
*unsupervised classification* problem [@figueiredo_unsupervised_2002].
The term *unsupervised* refers to the fact that the outcome variable 
-- true class membership -- is not known, 
and the term *classification* refers to the fact that 
the algorithm is predicting a categorical outcome -- class membership.

## A Taxonomy of Latent Class Analysis Methods

In this paper, we use the term latent class analysis to refer to 
a family of techniques that estimate latent class membership 
based on a parametric model of observed indicators.
From a historical perspective, the term was initially conceived to 
refer to analyses with categorical, usually binary indicators [@vermunt_jk_latent_2004].
Nowadays, there are a number of related techniques, 
known by distinct names, that serve a similar purpose.
The term "latent class analysis" seems most appropriate 
as an umbrella term for this broader class of models,
as it only refers to the purpose of the analysis, 
and does not imply restrictions to the model used, 
or the level of measurement of the indicators.
Given the abundance of terms in use for closely related classes of models, 
we will provide a rudimentary taxonomy of LCA methods.

A common type of LCA is the *finite Gaussian mixture model*; 
a univariate analysis where the observed distribution of a single variable 
is assumed to result from a mixture of a known number of normal (Gaussian) distributions.
The parameters of a finite Gaussian mixture model are 
the means and variances of these underlying normal distributions.
The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
In the multivariate case, with more than one indicator variable,
the parameters of a mixture model are the means, variances, and covariances 
between the indicators (which can be standardized to obtain correlations).
These parameters can be estimated freely, or set to be constrained across classes.

The technique known as *latent profile analysis (LPA)* is a special case of the mixture model,
which assumes conditional independence of the indicators.
Conditional independence means that, after class membership is accounted for,
the covariances/correlations between indicators are assumed to be zero.
This can be conceived of as a restricted mixture model with covariances fixed to zero.
In some cases, such constraints will be inappropriate;
for instance when the cohesion between indicators is expected to differ between classes.
As an example, a LPA of ocean plastic particles 
found two classes of particles based on length and width:
a class of smaller particles with a high correlation between length and width, 
meaning that these particles were approximately round or square in shape, 
and a class of larger particles with a low correlation between length and width, 
meaning that these particles were heterogenous in shape.
From a theoretical perspective, this makes sense, 
because the smaller particles have been ground down to a more uniform shape by the elements.
<!-- DA: what is the source for plastic particles? -->

It is also possible to estimate a mixture model based on latent indicators.
This means that, within each class, 
one or more continuous latent variables are estimated based on the observed indicators.
Categorical latent variable membership is then estimated 
based on these continuous latent variables.
A common application of this approach is in longitudinal research, 
where the indicators reflect one construct assessed at different time points.
Examples of this approach include *growth mixture models* (GMM) 
and *latent class growth analyses* (LCGA).
These techniques estimate a latent growth model
to describe individual trajectories over time.
The growth mixture model is a latent class model 
where the parameters that indicate class membership 
are the intercepts and variances, 
and typically covariances of the latent growth variables, 
e.g., a latent intercept and slope.
This technique assumes that individuals within a class can have heterogenous trajectories.
If the variance of the growth parameters is fixed to zero, 
it is known as a latent class growth analysis.
This latter approach assumes that all individuals within a class 
share the same identical trajectory,
and that any variance in the indicators not explained by the class-specific latent trajectories 
is due to residual error variance.

@@

This paper will primarily focus on mixture models and special cases thereof,
although most of the suggested guidelines are applicable to all LCA methods.

## Use Cases for Latent Class Analysis

<!-- Caspar will check if I can replace them with examples from Caspar's published papers. -->

There are several use cases for which LCA methods are suitable.
One example is to test a theory that postulates the existence of a categorical latent variable.
For example, *identity status theory* posits that, at any given point in time, 
adolescents reside in one of four identity statuses.
Latent class analysis can be used to identify these four statuses 
based on observed indicators (e.g., self-reported identity exploration and commitment).
If results indicate that the data are better described by a different number of classes,
or that the four-class solution does not correspond to the predicted pattern of responses on the indicators, 
then the theory may be called into question.

Another use case is unsupervised learning:
when the goal is to restore unobserved class membership based on observed indicators,
or to classify individuals.
For example, a mixture model can be used as a diagnostic aid 
when several clinical indicators can be used to distinguish between 
a fixed number of physical [@baughman_mixture_2006] or mental [@wu_abuse_2011] health problems.
The example of shoe size is a rudimentary illustration of this type of application.

@@

With LCA, our goal could be to inductively identify the number of classes. 
For instance, if we believe that a variable represents a group, 
but don't know how many groups there are, 
LCA may be an appropriate technique to answer this question.
For example, @hopfer_social_2014 studied substance use, sexual behavior, 
and mental health status of urban population in Winnipeg, Canada. 
The underlying assumption was that there were different risk profiles,
but their number was not known.
From a collection of indicators, LCA provided evidence 
that there are four distinct risk profiles in the Winnipeg area.

Another application of LCA is to classify individuals.
In a peer harassment study, 
@giang_using_2008 used LCA to classify over 2,000 sixth grade students 
into aggressor and victim latent classes.
The five-class solution comprised of classes of 
victims, aggressors, and socially adjusted students. 
For instance, it revealed that there were two types of victims: 
highly-victimized aggressive-victims and highly-aggressive aggressive-victims.

LCA is also appropriate when we wish to identify indicators that capture classes well.
High quality indicators will be strongly related to the latent variable
and will lead to good class separation.
This relationship of high quality indicators to the latent variable
is reflected in very high or very low conditional response probabilities.
For a simulation study exploring the effects of indicator quality on LCA, see @geiser_is_2014.
Therefore, one could use conditional response probabilities for each item to assess its quality
with regards to how well it helps separate the latent classes.
From this, a theory about the selection of indicators could be informed.

An extension of LCA is that containing covariates which can be used to predict class membership.
In this approach, we not only model the latent class variable based on indicators,
but we also relate the class membership to other explanatory variables (@vermunt_latent_2017).
An example of using covariates comes from @nozadi_moderating_2016 
who applied LCA to identify the probability of children's membership to an anxiety class.
The authors tested several covariates including children's age, sex, and accuracy scores.
Age and sex were not found to be related to the children's latent class membership,
hence these covariates were excluded from the analysis.
Accuracy scores, however, were related to probabilities of 
being in anxiety and attention-anxiety classes. 
Therefore this covariate was kept as a valuable predictor.

@@

<!-- CJ: (on the next paragraph)  -->
<!-- the multilevel LCA addresses violations of the assumptions of LCA. -->
<!-- That is not what this paragraph is about.  -->
<!-- This paragraph is about motivations for using LCA. -->
<!-- One motivation is if your data violates multivariate normality -->
<!-- because violating multivariate normality means that you have multiple groups -->
<!-- and you can use LCA to extract those. That is the only assumption  -->
<!-- we have to discuss here, we just need to explain it. -->
Finally, LCA can be used to deal with data which violate certain assumptions.
As discussed in the shoe size example, LCA can deal with violations of normality.
In fact, LCA assumes the population distribution is a non-normal mixture 
of $K$ normal distributions,
and it can discover the value of $K$, i.e. generate a $K$-class solution.

# Best Practices

## Best Practices in Estimation

The best practices in estimation, as outlined in Table \@ref(tab:checkest),
are rooted in existing recommendations for best practices 
for estimating specific subtypes of LCA,
including latent class growth analysis [@van_de_schoot_grolts-checklist_2017]
and latent class analysis with ordinal indicators [e.g., @nylund-gibson_ten_2018].
These were generalized to be relevant to all types of LCA,
and updated to current best practices, as explained below.

```{r checkest, include=TRUE}
tab_check <- data.frame(Item = c(
  "Examining Observed Data",
  "Handling Missing Data",
  "Alternative Model Specifications",
  "Software",
  "Algorithm",
  "Class Enumeration",
  "Model Fit Indices",
  "Classification Diagnostics",
  "Interpreting Class Solutions",
  "Label switching"

@@

### Examining Observed Data

Examining observed data is essential for any analysis
as it may reveal patterns and violations of assumptions 
that had not been considered prior to data collection.
Special attention should be paid 
to level of measurement of the indicators.
Finite Gaussian mixture models (including LPA) 
are only suitable for continuous variables.
Indicators with an ordinal level of measurement 
are likely to violate the assumption of 
within-class normal distributions of mixture models
[see @vermunt_k-means_2011].
Personal experience consulting on LCA methods and 
moderating the `tidyLPA` Google group
suggest that the misapplication of mixture models 
to ordinal (e.g., Likert-type) indicators 
is the most common source of user error.
Whereas it has been argued that some parametric methods are robust 
when scales with 7+ indicators are treated as continuous 
[e.g., @norman_likert_2010],
this certainly does not imply that all methods are.
It is certainly unlikely that such ordinal variables 
can be treated as a *mixture* of multiple normal distributions.
The problem becomes egregious when 
the number of classes estimated equals or exceeds the number of categories; 
in this case, each class-specific mean could describe a single response category, 
and a class-specific variance component would be nonsensical.
In sum, Likert-type scales are rarely suitable for mixture modeling;
latent class analysis with ordinal indicators is more appropriate.

Relatedly, a recent publication claimed that 
an assumption of mixture models is that 
observed indicators are normally distributed [@spurk_latent_2020].
This is incorrect.
When the number of classes is greater than one, 
mixture models assume that the observed indicators 
are a mixture of multiple (multivariate) normal distributions.
In our shoe size example, it can be seen that the population distribution
is comprised of two normal distributions.
When examined visually, the population distribution is evidently bimodal. 
The Shapiro-Wilk normality test ($W = 0.971$, $p < 0.05$ ) 
rejects the null hypothesis that 
the sample comes from a normally distributed population.
Yet, this is a prototypical example of a mixed population distribution 
where LCA can discover latent groups.
If the population distribution were instead normal, 
there would be no classes to extract 
as the whole population would belong to a single class.

Extensive descriptive statistics 
(including the number of unique values, 
variance of categorical variables, and missingness; see next subsection)
can be obtained using the function `tidySEM::descriptives(data)`.
Note, however, that sample-level descriptive statistics 
are of limited value when the goal of a study is 
to identify subsamples using latent class analysis.
Plots (density plots for continuous variables, 
and bar charts for categorical ones) may be more diagnostic.
Note that density plots can also aid in the choice of the number of classes, 
as further explained in the section on visualization.
Descriptive statistics and plots can be relegated to online supplements, 
provided that these are readily accessible 
[consider using a GitHub repository as a comprehensive public research archive, 
as explained in @van_lissa_worcs_2021].

### Handling Missing Data

Previous work has emphasized the importance of examining 
the pattern of missing data and 
reporting how missingness was handled 
[@van_de_schoot_grolts-checklist_2017].
Three types of missingness have been distinguished
in the literature [@rubin_inference_1976]:
Missing completely at random (MCAR), 
which means that missingness is random;
missing at random (MAR), 
which means that missingness is contingent on the *observed* data
(and can thus be accounted for);
and finally missing not at random (MNAR), 
which means that missingness is related to unobserved factors.
It is possible to conduct a so-called "MCAR" test, 
for example the non-parametric MCAR test [@jamshidian_tests_2010].
But note that the name "MCAR test" is somewhat misleading,
as the null-hypothesis of this test is that the data are not MAR,
and a significant test statistic indicates that 
missingness is related to the observed data (MAR).
A non-significant test statistic does not distinguish between MCAR or MNAR.
As Little's classic MCAR test relies on 
the comparison of variances across groups 
with different patterns of missing data, 
it assumes normality [@little_test_1988].
This assumption is tenuous in the context of LCA.
A non-parametric MCAR test, 
as provided by Jamshidian and Jalal, 
may be more suitable [@jamshidian_tests_2010].
Unfortunately, this test was removed from the central R-repository CRAN 
due to lack of maintenance.
For this tutorial, I have re-implemented it in the `mice` package 
as `mice::mcar()`,
with a fast backend in C++ and new printing and plotting methods.
